---
title: "figure1"
format: html
editor: visual
---

## Required Libraries

::: callout-note
All bash commands assumed executed from the root of this repository
:::

This notebook uses the following software and versions to pull data:

-   miniforge v24.7.1
-   amdirt=1.6.2
-   posieidon-xerxes=1.5.7.0
-   poseidon-trident=1.0.1.1

This can be recreated with

<!--
TODO: make a lock file?
-->

```{bash, eval=F}
conda env create -f figure1/environment.yaml
conda activate bergstrom2024
```

For data preparation and plotting

And the following libraries, and versions

```{r}
library(tidyverse)
library(ggsci)
sessionInfo()
```

## Getting Raw Data

This plot will initially show number of ancient metagenomes from host-associated contexts, ancient microbial genomes, and ancient human genome/genome-wide data.

Ancient environmental metagenomes are excluded as I believe AncientMetagenomeDir has a gross under-estimate of these, and I'm not sure if there is another authoritative source for this type of data.

For the (currently) three data types, we will just retrieve the number of samples as library information is not consistently recorded for human genome/genome-wide data.

### Ancient Host-Associated Metagenomes

Download latest ancient host-associated metagenome data from AncientMetagenomeDir using AMDirT,

```{bash, eval=F}
AMDirT download -t ancientsinglegenome-hostassociated -y samples -r v24.09.0 -o figure1/
# mv ancientmetagenome-hostassociated_samples_v24.09.0.tsv figure1/ -> apparently `-o` ignored?
```

Parse the file to extract just date and sample information; calculating cumulative sum in each year.

```{r}
data_raw_ancientmetagenome <- read_tsv('ancientmetagenome-hostassociated_samples_v24.09.0.tsv')

data_subset_ancientmetagenome <- data_raw_ancientmetagenome |> 
  select(publication_year) |> 
  group_by(publication_year) |> 
  summarise(
    n_samples_per = n(),
    sample_type = 'Ancient host-associated metagenome'
    ) |> 
  mutate(n_samples_cumul = cumsum(n_samples_per))
```

### Ancient Microbial Genomes

Download latest ancient host-associated singleagenome data from AncientMetagenomeDir using AMDirT

```{bash, eval=F}
AMDirT download -t ancientsinglegenome-hostassociated -y samples -r v24.09.0 -o figure1/
# mv ancientsinglegenome-hostassociated_samples_v24.09.0.tsv figure1/ -> apparently `-o` ignored?
```

Parse the file to extract just date and sample information; calculating cumulative sum in each year.

```{r}
data_raw_ancientmicrobegenome <- read_tsv('ancientsinglegenome-hostassociated_samples_v24.09.0.tsv')

data_subset_ancientsinglegenome <- data_raw_ancientsinglegenome |> 
  select(publication_year) |> 
  group_by(publication_year) |> 
  summarise(
    n_samples_per = n(),
    sample_type = 'Ancient microbial genome'
    ) |> 
  mutate(n_samples_cumul = cumsum(n_samples_per))
```

### Ancient Human Genomes

Pull summary statistics with Poseidon's trident

```{bash}
trident list --packages --remote --raw > figure1/ancienthumangenomewidedata.tsv
```

We couldn't do that because it isn't yet propogated with latest statistics.

We have to fall back to the AADR table and do alot of manual clean up.

Download from the AADR dataverse repo

```{bash, eval=F}
wget -O figure1/v62.0_1240k_public.anno https://dataverse.harvard.edu/api/access/datafile/10537413
```


First we remove modern human data, then we have perform a unique on the master ID,
which is the closest to a sample/individual identifier.
This is despite AADR themselves reporting 13571 unique individuals, I identified that there are many duplicate Master ID entries that seem to be derived from different library types (e.g. KPT002)

```{r}
data_raw_ancienthuman <- read_tsv('v62.0_1240k_public.anno')

data_subset_ancienthuman <- data_raw_ancienthuman |> 
  rename(
    id = `Master ID`, 
    publication_year = `Year data from this individual was first published [for a present-day individuals we give the data of the data reported here; missing GreenScience 2010 (Vi33.15, Vi33.26), Olalde2018 (I2657), RasmussenNature2010 (Australian)]`, 
    age = `Date mean in BP in years before 1950 CE [OxCal mu for a direct radiocarbon date, and average of range for a contextual date]`
  ) |> 
  select(
    id, publication_year, age
    ) |> 
  filter(age > 0) |> 
  unique() |> 
  select(id, publication_year) |> 
  group_by(publication_year) |> 
  summarise(
    n_samples_per = n(),
    sample_type = 'Ancient human individuals'
    ) |> 
  mutate(
    publication_year = as.numeric(publication_year),
    n_samples_cumul = cumsum(n_samples_per)
    )
```

## Non-human genomes

We also include non-human eukaryote genomes - namely plants and animals.

This comes from Bergström (pers. comm.).

```{r}
data_raw_ancientnonhuman <- read_tsv('nonhuman-published-genomes-AB151124.tsv')

## Not sure why by summarize plus cumsum wasn't working properly as above
## therefore went for a more convoluted strategy
data_subset_animal <- data_raw_ancientnonhuman |>
  filter(kingdom == 'Animal',
         year < 2025) |>
  select(year, nr_genomes) |>
  group_by(year) |>
  mutate(n_samples_per = sum(nr_genomes)) |>
  select(year, n_samples_per) |>
  unique() |>
  rename(publication_year = year, ) |>
  ungroup() |>
  select(publication_year, n_samples_per) |>
  arrange(publication_year) |>
  mutate(n_samples_cumul = cumsum(n_samples_per),
         sample_type = 'Ancient animal genome')

data_subset_plant <- data_raw_ancientnonhuman |>
  filter(kingdom == 'Plant',
         year < 2025) |>
  select(year, nr_genomes) |>
  group_by(year) |>
  mutate(n_samples_per = sum(nr_genomes)) |>
  select(year, n_samples_per) |>
  unique() |>
  rename(publication_year = year, ) |>
  ungroup() |>
  select(publication_year, n_samples_per) |>
  arrange(publication_year) |>
  mutate(n_samples_cumul = cumsum(n_samples_per),
         sample_type = 'Ancient plant genome')

data_subset_fungi <- data_raw_ancientnonhuman |>
  filter(kingdom == 'Fungi',
         year < 2025) |>
  select(year, nr_genomes) |>
  group_by(year) |>
  mutate(n_samples_per = sum(nr_genomes)) |>
  select(year, n_samples_per) |>
  unique() |>
  rename(publication_year = year, ) |>
  ungroup() |>
  select(publication_year, n_samples_per) |>
  arrange(publication_year) |>
  mutate(n_samples_cumul = cumsum(n_samples_per),
         sample_type = 'Ancient fungi genome')

```

### Combine Data

Combine the different datasets, and get some basic additional information for better plotting readability.

```{r}
data_subset_combined <- data_subset_ancientmetagenome |>
  bind_rows(
    list(
      data_subset_ancienthuman,
      data_subset_ancientsinglegenome,
      data_subset_animal,
      data_subset_plant,
      data_subset_fungi
    )
  ) |>
  arrange(publication_year)

data_year_range = seq(
  min(data_subset_combined$publication_year),
  max(data_subset_combined$publication_year),
)

data_type_order <- data_subset_combined |>
  select(sample_type, n_samples_cumul) |> 
  group_by(sample_type) |> 
  summarise(current = max(n_samples_cumul)) |> 
  arrange(current) |> 
  mutate(sample_type = as_factor(sample_type)) |> 
  pull(sample_type)
```

Plot a stacked bar plot

```{r}
ggplot(data_subset_combined |> mutate(sample_type = factor(sample_type, levels = data_type_order)), aes(publication_year, n_samples_cumul, fill = sample_type)) +
  geom_col() +
  scale_fill_npg(breaks = data_type_order) +
  theme_classic() +
  theme(
    axis.line.y = element_blank(),
    legend.position = "top"
  ) +
  xlab('') +
  ylab('Number of samples with sequencing data') + 
  guides(fill = guide_legend(title = ""))
```

